---
title: "Computational Methods 1 - Assignment 1"
author: "Luke Korthals"
date: "2024-05-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Imports, global settings, and Functions

I am using `tidyverse` for data manipulation and plotting. To make sure plots are consistent, a color set is defined using `RColorBrewer`.
```{r, echo=TRUE, message=FALSE, warning=FALSE}	
# Imports
library(tidyverse)

# Global settings
plot_colors <- RColorBrewer::brewer.pal(8, "Set2")
```

The following functions are used throughout the assignment.
```{r, echo=TRUE, message=FALSE, warning=FALSE}	
# Functions
prob_cw <- function(x, lambda = 0.5) {
    # Computes the Curie-Weiss likelihood for a given data vector x and lambda
    # returns: the likelihood
    # ----------
    # x: data vector
    # lambda: parameter of the Curie-Weiss likelihood

    if (is.null(dim(x))) {
        m <- 1
        n <- length(x)
    }
    if (!is.null(dim(x))) {
        m <- dim(x)[1]
        n <- dim(x)[2]
    }

    pr_cw_unnorm <- function(n, r, lambda) choose(n, r) * exp(-lambda * r * (r - 1))
    pr_cw_i <- array(NA, dim = c(m, length(lambda)))
    pr_cw <- c()
    Z_cw <- 0
    for (i in 0:n) Z_cw <- Z_cw + pr_cw_unnorm(n, i, lambda)
    for (k in 1:m) {
        if (!is.null(dim(x))) r <- sum(x[k, ])
        if (is.null(dim(x))) r <- sum(x)
        pr_cw_i[k, ] <- pr_cw_unnorm(n, r, lambda) / Z_cw
    }
    for (j in 1:length(lambda)) pr_cw[j] <- prod(pr_cw_i[, j])
    return(pr_cw)
}


plot_gamma_ <- function(s, r, name, stat_fun_only = FALSE) {
    # Plots a gamma distribution with shape s and rate r
    # returns: Either a ggplot object or a stat_function object that can be added to a ggplot object
    # ----------
    # s: shape parameter of the gamma distribution
    # r: rate parameter of the gamma distribution
    # name: name of the distribution
    # stat_fun_only: if TRUE, returns a stat_function object only

    # Compute Gamma distribution
    ps <- stat_function(fun = dgamma, args = list(shape = s, rate = r), aes(color = name), linewidth = 1)
    if (stat_fun_only) {
        return(ps)
    }

    # Create plot
    xmin <- qgamma(1e-25, s, r)
    xmax <- qgamma(0.99999, s, r)
    p <- data.frame(x = c(xmin, xmax)) %>%
        ggplot(aes(x = x)) +
        ps +
        labs(x = expression(lambda), y = "Density") +
        scale_color_manual(values = plot_colors, name = "Distribution") +
        theme_classic() +
        coord_cartesian(expand = TRUE, clip = "off")

    return(p)
}

gamma_cw_grid_approx <- function(s, r, x, lambda_grid, n_samples) {
    # Performs grid approximation using a gamma prior and Curie-Weiss likelihood
    # returns: n_samples samples from the posterior
    # ----------
    # s: shape parameter of the gamma prior
    # r: rate parameter of the gamma prior
    # x: data vector
    # lambda_grid: grid of values for lambda
    # n_samples: number of samples to draw from the posterior

    dat <- data.frame(lambda = lambda_grid) # Create grid
    dat <- dat %>%
        mutate(
            prior = dgamma(lambda, s, r), # Evaluate prior
            likelihood = prob_cw(x, lambda), # Evaluate likelihood
            unnormalized_posterior = prior * likelihood, # Calculate unnormalized posterior
            posterior = unnormalized_posterior / sum(unnormalized_posterior) # Normalized posterior
        )

    return(sample_n(dat, n_samples, replace = TRUE, weight = posterior)) # Sample from posterior
}
```



### Question 1

> From Chapter 6 of Bayes Rules (BR) we obtain a method in four steps to apply the grid approximation to the posterior. We use this grid approximation for $\lambda$ here. 
We use the <span style="color:orange">Gamma prior with parameters $s = 2$ and $r = 2$</span>. 
Use <span style="color:orange">grid approximation for the posterior $f(\lambda | x1, x2)$</span>. 
We have the <span style="color:orange"> data $x1 = 1$ and $x2 = 1$</span>.

First, the parameter for the Gamma prior and the data are defined.
```{r}
# Prior
s <- 2
r <- 2

# Data
x1 <- 1
x2 <- 1
```	


> (a) Make a <span style="color:orange">plot of the Gamma distribution</span> with parameters s = 2 and r = 2.
```{r}	
plot_gamma_(s, r, "a: Prior")
```

As can be seen from this plot, the prior assumes that the weight of the edge $\lambda$ likely lies between 0 and 2.


> (b) Define a <span style="color:orange">grid for λ with values {0, 0.5, 1, 1.5, 2}</span> and create the grid data structure for the parameter lambda grid, as shown in Chapter 6.**
```{r}	
lambda <- seq(0, 2, length = 5)
dat <- data.frame(lambda)
head(dat)
```	


> (c) <span style="color:orange">Add to grid data the prior and the likelihood.</span> 
The likelihood you can add by applying the function prob cw in the R file on canvas. 
The function has two arguments, the data vector (x1, x2) and the value (or values for the grid) for λ.
```{r}
dat <- dat %>%
    mutate(
        prior = dgamma(lambda, s, r),
        likelihood = prob_cw(c(x1, x2), lambda)
    )
head(dat)
```
The data above shows that likelihood $L(\lambda|{1, 1})$ is highest for $\lambda = 0$ and $\lambda = 1$.

>(d) <span style="color:orange">Add to the structure grid data the unnormalized posterior.</span> 
Then in the final step use the sample n function as shown in the book BR to <span style="color:orange">obtain 10,000 posterior samples</span>. 
And <span style="color:orange">plot the results</span>, both for the prior and for the posterior.
```{r}
# Calculate posterior
dat <- dat %>%
    mutate(
        unnormalized_posterior = prior * likelihood,
        posterior = unnormalized_posterior / sum(unnormalized_posterior)
    )
head(dat)
```
The data above shows that the posterior probability $f(\lambda|{x1, x2})$ is highest for $\lambda = 1$. 
Even though $\lambda = 0$ had the highest likelihood, its prior probability $P(\lambda = 0) = 0$, making the corresponding posterior probability $P(\lambda = 0|{x1, x2}) = 0$ as well.
```{r}
# Simulate posterior
n_samples <- 10000
simulated_posterior_5 <- sample_n(dat, n_samples, replace = TRUE, weight = posterior)

# Plot
plot_gamma_(s, r, "a: Prior") +
    geom_histogram(
        data = simulated_posterior_5, 
        aes(x = lambda, color = "b: Simulated Posterior (grid size 5)", after_stat(density)), 
        fill = plot_colors[2], 
        alpha = 0.5
        )
```	
This plot shows that the simulated posterior is quite rough which suggests that the grid size is too small. 

> (e) <span style="color:orange">Repeat the analysis</span> in (a)-(d) but now use a <span style="color:orange">grid of 501 equally spaced values between 0 and 5 for λ</span>.
```{r}
# Define grid
lambda <- seq(0, 5, length = 501)

# Simulate posterior
simulated_posterior_501 <- gamma_cw_grid_approx(s, r, c(x1, x2), lambda, n_samples) # Function defined above

# Plot
plot_gamma_(s, r, "a: Prior") +
    geom_histogram(
        data = simulated_posterior_501, 
        aes(x = lambda, color = "b: Simulated Posterior (grid size 501)", after_stat(density)), 
        fill = plot_colors[2], 
        alpha = 0.5
        )
```	

The plot above shoes that the approximation using a grid size of 501 is much smoother than the approximation using a grid size of 5.


### Question 2

> We will take a closer look at the combination of the likelihood and the Gamma prior here. 
In Section 5.2.4 of BR it is shown that the Poisson probability for the likelihood is conjugate to the Gamma prior. 
<span style="color:orange">Here we will show that the combination of the Curie-Weiss likelihood and the Gamma prior are conjugate</span> for the unnormalised likelihood.


> (a) Ignoring the normalising constants for the Gamma prior and the likelihood prior, <span style="color:orange">show in a similar way as in BR, that the posterior with the Gamma prior is conjugate to the Curie-Weiss model</span> given above.**

Our prior $f(\lambda)$ is given by a Gamma pdf, the likelihood $L(\lambda | x1, x2)$ is given by the Curie-Weiss model, and the posterior $f(\lambda | x1, x2)$ is proportional to the product of the prior and the likelihood.
\[
  f(\lambda) = \frac{r^s}{\Gamma(s)}\lambda^{s-1}e^{-r\lambda} \\
  L(\lambda | {x1, x2}) = \frac{1}{Z_{CW}}e^{-\lambda(x1+x2)(x1+x2-1)} \\
  f(\lambda | {x1, x2}) \propto \lambda^{s-1}e^{-r\lambda} \cdot e^{-\lambda(x1+x2)(x1+x2-1)} = \lambda^{s-1}e^{\lambda(-r)-(x1+x2-1)\lambda(x1+x2)}
\]

Any constants that do not depend on $\lambda$ can be dropped and the remaining kernel is still proportional to the posterior.
\[
  f(\lambda | {x1, x2}) \propto \lambda^{s-1}e^{-r\lambda} \cdot e^{-\lambda(x1+x2)(x1+x2-1)} = \lambda^{s-1}e^{\lambda(-r)-(x1+x2-1)\lambda(x1+x2)}
\]

This kernel corresponds to a Gamma pdf, thus, the resulting posterior is conjugate to the prior Gamma.
\[
  \lambda | {x1, x2} \sim Gamma({s, r+(x1+x2-1)(x1+x2)})
\]



> (b) What are the <span style="color:orange">parameters of the posterior distribution</span> of the pair Gamma-Curie-Weiss?**

In Our case, we have: 
\[
    s = r = 2 \\
    x1 = x2 = 1 \\ 
    f(\lambda) \sim Gamma({2, 2}) \\ 
    \lambda | {1, 1} \sim Gamma({2, 2+(1+1-1)(1+1)}) = Gamma({2, 4})
\]



> (c) <span style="color:orange">Plot the prior, the sampled posterior and the posterior</span> in a single plot. 
Use the prior Gamma(2, 2), the data x1 = 1 and x2 = 1.**
```{r}
# Calculate new Gamma parameters
s_new <- s
r_new <- r + (x1 + x2 - 1) * (x1 + x2)

# Plot
plot_gamma_(s, r, "a: Prior") +
    geom_histogram(data = simulated_posterior_501, aes(x = lambda, color = "b: Simulated Posterior (grid size 501)", after_stat(density)), fill = plot_colors[2], alpha = 0.5) +
    plot_gamma_(s_new, r_new, "c: Posterior", stat_fun_only = TRUE)
```

This plot shows that the simulated posterior with a grid size of 501 approximates the true posterior quite well. 
